---
title: "Assessing Matching Quality"
author: "Daniel Ahn"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

Purpose of this RMD is to determine the quality of our exact matching schemes. For this, we used the cobalt package in R. First, we created Love Plots that visualize the Kolmogorov-Smirnov statistics as well as standardized mean differences pre- and post-exact matching. For both laboratory and allocation MELD/PELD scores, our statistics were 0 after matching, indicating identical distributions. These distributions can actually be seen in our balance plots, in the second chunk below.

```{r matching_quality_statistics}
v <- data.frame(old = c("Allocation_Score", "PELD_or_MELD", "Laboratory_Score"),
                new = c("Allocation Score", "PELD or MELD", "Laboratory Score"))

love.plot(m.out, stats = c("mean.diffs", "ks.statistics"), 
          binary = "std", abs = TRUE,
          var.order = "unadjusted", var.names = v,
          limits = c(0, 3.0), grid = FALSE, wrap = 10,
          sample.names = c("Unmatched", "Matched"),
          position = "top", shapes = c("circle", "triangle"),
          colors = c("red", "blue")) 

love.plot(m.out1, stats = c("mean.diffs", "ks.statistics"), 
          binary = "std", abs = TRUE,
          var.order = "unadjusted", var.names = v,
          limits = c(0, 1.0), grid = FALSE, wrap = 10,
          sample.names = c("Unmatched", "Matched"),
          position = "top", shapes = c("circle", "triangle"),
          colors = c("green", "orange")) 

```

```{r matching_quality_distributions}

match1 <- bal.plot(m.out, var.name = "Allocation_Score", which = "both",
          type = "density", mirror = TRUE) + 
          theme(legend.position = "none") +
          scale_fill_manual(values = c("#ddd3bb", "#e49f62")) +
          scale_x_continuous(name = "Allocation Score") +
          ggtitle("Distribution Pre/Post-Exact Matching by Allocation Score")

match2 <- bal.plot(m.out, var.name = "PELD_or_MELD", which = "both",
          type = "density", mirror = TRUE) + 
          scale_fill_manual(name = "Exception Status", values = c("#ddd3bb", "#e49f62"), labels = c("Non-Exception", "Received Exception")) +
          scale_x_discrete(name = "PELD or MELD", labels = c("PELD", "MELD"))  +
          ggtitle("Distribution Pre/Post-Exact Matching by PELD or MELD")

match3 <- bal.plot(m.out1, var.name = "Laboratory_Score", which = "both",
          type = "density", mirror = TRUE) + 
          theme(legend.position = "none") +
          scale_fill_manual(values = c("#496061", "#8dd9d7")) +
          scale_x_continuous(name = "Laboratory Score") +
          ggtitle("Distribution Pre/Post-Exact Matching by Laboratory Score")


match4 <- bal.plot(m.out1, var.name = "PELD_or_MELD", which = "both",
          type = "density", mirror = TRUE) + 
          scale_fill_manual(name = "Exception Status", values = c("#496061", "#8dd9d7"), labels = c("Non-Exception", "Received Exception")) +
          scale_x_discrete(name = "PELD or MELD", labels = c("PELD", "MELD")) +
          ggtitle("Distribution Pre/Post-Exact Matching by PELD or MELD")

plot_grid(match1, match2,
          ncol = 2, nrow = 1,
          labels = c("A", "B"))

plot_grid(match3, match4,
          ncol = 2, nrow = 1,
          labels = c("A", "B"))


```




