---
title: "Mixed Effects Cox Models"
author: "Daniel Ahn"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

Here, I load in three files to set up the data for the mixed effects models. First is p1, which is the dataset of 2026 candidates. Next is stathist_liin, which has all status updates during candidates' time spent on the waitlist. This information includes whether patients had applied for a MELD/PELD exception. Last is mpexcept, which presents information on the status of the MELD/PELD exception applications (approval, denial, withdrawal, etc.) as well as the time periods when these exceptions are active. All three files are combined to set up a data file amenable for time-varying covariates of interest, including active non-standardized exception status as well as allocation MELD/PELD score. 

```{r model_setup}

stathist <- stathist_liin %>% dplyr::select(c("PX_ID", "CANHX_BEGIN_DT", "CANHX_END_DT", 
                                              "CANHX_EXC_FLG", "CANHX_EXC_SCORE",
                                              "CANHX_OPTN_LAB_MELD", "CANHX_SRTR_LAB_MELD",
                                              "CANHX_STAT_CD", "CANHX_BILI", "CANHX_INR",
                                              "CANHX_SERUM_CREAT", "CANHX_SERUM_SODIUM", "CANHX_DIAL_PRIOR_WEEK",
                                              "CAN_LAST_ACT_STAT_DT"))

mpexceptshort2 <- dplyr::select(mpexcept, c(
  "PX_ID",
  "CANHX_MPXCPT_APPL_TY",
  "CANHX_MPXCPT_BEGIN_DT",
  "CANHX_MPXCPT_CASE_CUR_FORM",
  "CANHX_MPXCPT_CASE_ID",
  "CANHX_MPXCPT_DGN",
  "CANHX_MPXCPT_END_DT",
  "CANHX_MPXCPT_FORM_ID",
  "CANHX_MPXCPT_MELD_OR_PELD",
  "CANHX_MPXCPT_MOST_RECENT_APPROV",
  "CANHX_MPXCPT_ORIG_APPL_DT",
  "CANHX_MPXCPT_REQUEST_SCORE",
  "CANHX_MPXCPT_STAT",
  "CANHX_MPXCPT_STAT_DT"
))

forcox <- merge(p1, stathist, by = "PX_ID", all.x = TRUE)
forcox <- merge(forcox, mpexceptshort2, by = "PX_ID", all.x = TRUE) 

cox1 <- forcox %>% dplyr::select(c("PX_ID", "CAN_LISTING_DT", "CAN_REM_DT", "CAN_REM_CD", "POLICY_COHORT",
"CANHX_BEGIN_DT.x", "CANHX_END_DT.x", "CANHX_BEGIN_DT.y", "CANHX_END_DT.y", "SCORE", "LAB_SCORE", 
"SCORE_DIFFERENCE", "REC_EXCEPTION", "DEATH", "DEATH_DT", "CAN_LISTING_CTR_ID", "CANHX_EXC_FLG.y", "CANHX_EXC_SCORE.y", "CANHX_MPXCPT_APPL_TY.y", "CANHX_MPXCPT_BEGIN_DT.y", "CANHX_MPXCPT_DGN.y", "CANHX_MPXCPT_END_DT.y", "CANHX_MPXCPT_ORIG_APPL_DT.y", "CANHX_MPXCPT_REQUEST_SCORE.y", "CANHX_MPXCPT_STAT.y", "CANHX_MPXCPT_STAT_DT.y", "CANHX_OPTN_LAB_MELD.y", "CANHX_SRTR_LAB_MELD.y", "CANHX_STAT_CD.y", "CAN_GENDER", "CAN_RACE", "CAN_AGE_AT_LISTING", "CAN_AGE_IN_MONTHS_AT_LISTING", "CANHX_BILI", "CANHX_INR", "CANHX_SERUM_CREAT", "CANHX_SERUM_SODIUM", "CANHX_DIAL_PRIOR_WEEK", "CAN_LAST_ACT_STAT_DT.y")) %>%
  group_by(PX_ID) %>% filter(any(CANHX_EXC_FLG.y == 1)) %>% filter(CANHX_EXC_FLG.y == 1) %>% 
  filter(CANHX_MPXCPT_STAT.y == 5 | CANHX_MPXCPT_STAT.y == 14 | CANHX_MPXCPT_STAT.y == 16 | CANHX_MPXCPT_STAT.y == 18 | CANHX_MPXCPT_STAT.y == 30) %>%
  ungroup() %>% group_by(PX_ID, CANHX_BEGIN_DT.y) %>% arrange(PX_ID, CANHX_BEGIN_DT.y) %>%
  filter((CANHX_BEGIN_DT.y >= CANHX_MPXCPT_BEGIN_DT.y & CANHX_END_DT.y <= CANHX_MPXCPT_END_DT.y) |
           abs(as.numeric(difftime(CANHX_BEGIN_DT.y, CANHX_MPXCPT_BEGIN_DT.y, units = "days"))) <= 5 |
           abs(as.numeric(difftime(CANHX_END_DT.y, CANHX_MPXCPT_END_DT.y, units = "days"))) <= 5) %>%
  mutate(ACTIVE = 1,
         EXCEPTION = 1,
         NEW_SCORE = CANHX_EXC_SCORE.y - 6200,
         NEW_SCORE_DIFFERENCE = NEW_SCORE - (CANHX_SRTR_LAB_MELD.y - 6200))

cox1.1 <- forcox3.0 %>% ungroup() %>% group_by(PX_ID) %>% arrange(CANHX_BEGIN_DT.y) %>% slice_tail(n = 1) %>% filter(DEATH == 1) %>% 
  filter(abs(as.numeric(difftime(DEATH_DT, CANHX_END_DT.y, units = "days"))) != 0) %>%
  mutate(CANHX_BEGIN_DT.y = CANHX_END_DT.y + 1) %>% mutate(CANHX_END_DT.y = DEATH_DT)

cox2 <- forcox %>% dplyr::select(c("PX_ID", "CAN_LISTING_DT", "CAN_REM_DT", "CAN_REM_CD", "POLICY_COHORT",
"CANHX_BEGIN_DT.x", "CANHX_END_DT.x", "CANHX_BEGIN_DT.y", "CANHX_END_DT.y", "SCORE", "LAB_SCORE", 
"SCORE_DIFFERENCE", "REC_EXCEPTION", "DEATH", "DEATH_DT", "CAN_LISTING_CTR_ID", "CANHX_EXC_FLG.y", "CANHX_EXC_SCORE.y", "CANHX_MPXCPT_APPL_TY.y", "CANHX_MPXCPT_BEGIN_DT.y", "CANHX_MPXCPT_DGN.y", "CANHX_MPXCPT_END_DT.y", "CANHX_MPXCPT_ORIG_APPL_DT.y", "CANHX_MPXCPT_REQUEST_SCORE.y", "CANHX_MPXCPT_STAT.y", "CANHX_MPXCPT_STAT_DT.y", "CANHX_OPTN_LAB_MELD.y", "CANHX_SRTR_LAB_MELD.y", "CANHX_STAT_CD.y", "CAN_GENDER", "CAN_RACE", "CAN_AGE_AT_LISTING", "CAN_AGE_IN_MONTHS_AT_LISTING", "CANHX_BILI", "CANHX_INR", "CANHX_SERUM_CREAT", "CANHX_SERUM_SODIUM", "CANHX_DIAL_PRIOR_WEEK", "CAN_LAST_ACT_STAT_DT.y")) %>%
  group_by(PX_ID) %>% filter(any(CANHX_EXC_FLG.y == 1)) %>% filter(CANHX_EXC_FLG.y == 1) %>%
  ungroup() %>% group_by(PX_ID, CANHX_BEGIN_DT.y) %>% arrange(PX_ID, CANHX_BEGIN_DT.y) %>%
  slice_head(n = 1) %>% mutate(ACTIVE = 0,
                               EXCEPTION = 0,
                               NEW_SCORE = CANHX_SRTR_LAB_MELD.y - 6200,
                               NEW_SCORE_DIFFERENCE = NEW_SCORE - (CANHX_SRTR_LAB_MELD.y - 6200))

cox2.1 <- forcox3.1 %>% ungroup() %>% group_by(PX_ID) %>% arrange(CANHX_BEGIN_DT.y) %>% slice_tail(n = 1) %>% filter(DEATH == 1) %>% 
  filter(abs(as.numeric(difftime(DEATH_DT, CANHX_END_DT.y, units = "days"))) != 0) %>%
  mutate(CANHX_BEGIN_DT.y = CANHX_END_DT.y + 1) %>% mutate(CANHX_END_DT.y = DEATH_DT)

cox3 <- forcox %>% dplyr::select(c("PX_ID", "CAN_LISTING_DT", "CAN_REM_DT", "CAN_REM_CD", "POLICY_COHORT",
"CANHX_BEGIN_DT.x", "CANHX_END_DT.x", "CANHX_BEGIN_DT.y", "CANHX_END_DT.y", "SCORE", "LAB_SCORE", 
"SCORE_DIFFERENCE", "REC_EXCEPTION", "DEATH", "DEATH_DT", "CAN_LISTING_CTR_ID", "CANHX_EXC_FLG.y", "CANHX_EXC_SCORE.y", "CANHX_MPXCPT_APPL_TY.y", "CANHX_MPXCPT_BEGIN_DT.y", "CANHX_MPXCPT_DGN.y", "CANHX_MPXCPT_END_DT.y", "CANHX_MPXCPT_ORIG_APPL_DT.y", "CANHX_MPXCPT_REQUEST_SCORE.y", "CANHX_MPXCPT_STAT.y", "CANHX_MPXCPT_STAT_DT.y", "CANHX_OPTN_LAB_MELD.y", "CANHX_SRTR_LAB_MELD.y", "CANHX_STAT_CD.y", "CAN_GENDER", "CAN_RACE", "CAN_AGE_AT_LISTING", "CAN_AGE_IN_MONTHS_AT_LISTING", "CANHX_BILI", "CANHX_INR", "CANHX_SERUM_CREAT", "CANHX_SERUM_SODIUM", "CANHX_DIAL_PRIOR_WEEK", "CAN_LAST_ACT_STAT_DT.y")) %>%
  group_by(PX_ID) %>% filter(any(CANHX_EXC_FLG.y == 1)) %>% filter(CANHX_EXC_FLG.y == 0) %>% 
  ungroup() %>% group_by(PX_ID, CANHX_BEGIN_DT.y) %>% arrange(PX_ID, CANHX_BEGIN_DT.y) %>%
  slice_head(n = 1) %>% mutate(ACTIVE = 0,
                               EXCEPTION = 0,
                               NEW_SCORE = CANHX_SRTR_LAB_MELD.y - 6200,
                               NEW_SCORE_DIFFERENCE = NEW_SCORE - (CANHX_SRTR_LAB_MELD.y - 6200))

cox3.1 <- forcox3.2 %>% ungroup() %>% group_by(PX_ID) %>% arrange(CANHX_BEGIN_DT.y) %>% slice_tail(n = 1) %>% filter(DEATH == 1) %>% 
  filter(abs(as.numeric(difftime(DEATH_DT, CANHX_END_DT.y, units = "days"))) != 0) %>%
  mutate(CANHX_BEGIN_DT.y = CANHX_END_DT.y + 1) %>% mutate(CANHX_END_DT.y = DEATH_DT)

cox4 <- do.call("rbind", list(cox1, cox1.1, cox2, cox2.1, cox3, cox3.1)) %>%
  ungroup() %>% group_by(PX_ID, CANHX_BEGIN_DT.y) %>% arrange(ACTIVE) %>% slice_tail(n = 1)

cox5 <- forcox %>% dplyr::select(c("PX_ID", "CAN_LISTING_DT", "CAN_REM_DT", "CAN_REM_CD", "POLICY_COHORT",
"CANHX_BEGIN_DT.x", "CANHX_END_DT.x", "CANHX_BEGIN_DT.y", "CANHX_END_DT.y", "SCORE", "LAB_SCORE", 
"SCORE_DIFFERENCE", "REC_EXCEPTION", "DEATH", "DEATH_DT", "CAN_LISTING_CTR_ID", "CANHX_EXC_FLG.y", "CANHX_EXC_SCORE.y", "CANHX_MPXCPT_APPL_TY.y", "CANHX_MPXCPT_BEGIN_DT.y", "CANHX_MPXCPT_DGN.y", "CANHX_MPXCPT_END_DT.y", "CANHX_MPXCPT_ORIG_APPL_DT.y", "CANHX_MPXCPT_REQUEST_SCORE.y", "CANHX_MPXCPT_STAT.y", "CANHX_MPXCPT_STAT_DT.y", "CANHX_OPTN_LAB_MELD.y", "CANHX_SRTR_LAB_MELD.y", "CANHX_STAT_CD.y", "CAN_GENDER", "CAN_RACE", "CAN_AGE_AT_LISTING", "CAN_AGE_IN_MONTHS_AT_LISTING", "CANHX_BILI", "CANHX_INR", "CANHX_SERUM_CREAT", "CANHX_SERUM_SODIUM", "CANHX_DIAL_PRIOR_WEEK", "CAN_LAST_ACT_STAT_DT.y")) %>% 
  group_by(PX_ID) %>% filter(all(CANHX_EXC_FLG.y == 0)) %>% ungroup() %>% group_by(PX_ID, CANHX_BEGIN_DT.y) %>%
  arrange(PX_ID, CANHX_BEGIN_DT.y) %>% slice_tail(n = 1) %>%
  mutate(
    ACTIVE = 0,
    EXCEPTION = case_when(
      !is.na(CANHX_EXC_SCORE.y) ~ 1,
      TRUE ~ 0),
    NEW_SCORE = case_when(
      EXCEPTION == 1 ~ CANHX_EXC_SCORE.y - 6200,
      TRUE ~ CANHX_SRTR_LAB_MELD.y - 6200),
    NEW_SCORE_DIFFERENCE = NEW_SCORE - (CANHX_SRTR_LAB_MELD.y - 6200)) 

cox5.1 <- cox5 %>% ungroup() %>% group_by(PX_ID) %>% arrange(CANHX_BEGIN_DT.y) %>% slice_tail(n = 1) %>% filter(DEATH == 1) %>% 
  filter(abs(as.numeric(difftime(DEATH_DT, CANHX_END_DT.y, units = "days"))) != 0) %>%
  mutate(CANHX_BEGIN_DT.y = CANHX_END_DT.y + 1) %>% mutate(CANHX_END_DT.y = DEATH_DT)

cox6 <- do.call("rbind", list(cox4, cox5, cox5.1)) %>%
  mutate(NEW_DEATH = case_when(
    DEATH == 1 & abs(as.numeric(difftime(DEATH_DT, CANHX_END_DT.y, units = "days"))) == 0 ~ 1,
    TRUE ~ 0))

forcoxfinal <- new %>% group_by(PX_ID) %>% arrange(CANHX_BEGIN_DT.y) %>%
  mutate(TIME1 = as.numeric(difftime(CANHX_BEGIN_DT.y, CAN_LISTING_DT, units = "days")),
         TIME2 = as.numeric(difftime(CANHX_END_DT.y, CAN_LISTING_DT, units = "days")) + 1,
         TIME2 = case_when(
           is.na(TIME2) ~ as.numeric(difftime("2021-12-31", CAN_LISTING_DT, units = "days")),
           TRUE ~ TIME2),
         SCORE_DIFF_GROUP = case_when(
           NEW_SCORE_DIFFERENCE <= 0 ~ "0",
           NEW_SCORE_DIFFERENCE > 0 & NEW_SCORE_DIFFERENCE <= 10 ~ "1-10",
           NEW_SCORE_DIFFERENCE > 10 & NEW_SCORE_DIFFERENCE <= 20 ~ "11-20",
           NEW_SCORE_DIFFERENCE > 20 & NEW_SCORE_DIFFERENCE <= 30 ~ "21-30",
           NEW_SCORE_DIFFERENCE > 30 ~ ">30"),
         SCORE_DIFF_GROUP = factor(SCORE_DIFF_GROUP, levels = c("0", "1-10", "11-20", "21-30", ">30")),
         sex = ifelse(CAN_GENDER == "M", "Male", "Female"),
         sex = factor(sex, levels = c("Male", "Female")),
         race = case_when(
           CAN_RACE == 8 ~ "White",
           CAN_RACE == 16 ~ "Black",
           CAN_RACE == 2000 ~ "Hispanic/Latino",
           CAN_RACE == 64 ~ "Asian",
           TRUE ~ "Other"),
         race = factor(race, levels = c("White", "Black", "Hispanic/Latino", "Asian", "Other")),
         age = CAN_AGE_AT_LISTING,
         agegroup = case_when(
           CAN_AGE_AT_LISTING < 1 ~ "<1",
           CAN_AGE_AT_LISTING >= 1 & CAN_AGE_AT_LISTING <= 5 ~ "1-5",
           CAN_AGE_AT_LISTING > 5 & CAN_AGE_AT_LISTING <= 11 ~ "6-11",
           CAN_AGE_AT_LISTING > 11 ~ "12-17"),
         agegroup = factor(agegroup, levels = c("<1", "1-5", "6-11", "12-17"))) 


```

Purpose of model 1 is to identify hazard ratio associated with active non-standardized exception status.
```{r model1}

model1 <- coxme(Surv(TIME1, TIME2, NEW_DEATH) ~ NEW_SCORE + EXCEPTION + race + sex + agegroup + (1 | CAN_LISTING_CTR_ID), data = forcoxfinal)
tab_model(model1)


```

Purpose of model 2 is to identify hazard ratio associated with every point increase that a patient receives over their laboratory MELD/PELD score due to an active non-standardized exception.
```{r model2}
model2 <- coxme(Surv(TIME1, TIME2, NEW_DEATH) ~ NEW_SCORE_DIFFERENCE + NEW_SCORE + race + sex + agegroup + (1 | CAN_LISTING_CTR_ID), data = forcoxfinal)
tab_model(model2)

```

Here, we ran model 1 again, but stratified by policy cohort to determine whether non-standardized exceptions were still associated with a significant difference in mortality risk after the NLRB's implementation.
```{r pre_and_post_cox}

coxpostpolicy <- filter(forcoxfinal, POLICY_COHORT == "Post-Policy")
coxprepolicy <- filter(forcoxfinal, POLICY_COHORT == "Pre-Policy")

premodel <- coxme(Surv(TIME1, TIME2, NEW_DEATH) ~ NEW_SCORE + EXCEPTION + race + sex + agegroup + (1 | CAN_LISTING_CTR_ID), data = coxprepolicy)
summary(premodel)

postmodel <- coxme(Surv(TIME1, TIME2, NEW_DEATH) ~ NEW_SCORE + EXCEPTION + race + sex + agegroup + (1 | CAN_LISTING_CTR_ID), data = coxpostpolicy)
summary(postmodel)

```
