---
title: "Figures"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

This RMD contains the code for the figures 2, 3, and 5 included in the main manuscript text. 

This code is for making custom functions for color schemes that will be used in the figures later.
```{r color_schemes}
custom_colors <- c(
  `dark red` = "#660000",
  `bright red` = "#b81d2e",
  `orange` = "#ae6320",
  `yellow`= "#b8a370",
  `green` = "#3ba9a9",
  `white` = "#FFFFFF",
  `grey` = "#C3C1C1",
  `light green` = "#8dd9d7",
  `light yellow` = "#ddd3bb",
  `light orange` = "#e49f62",
  `space sparkle`= "#496061",
  `violet` = "#7A3DE3",
  `sandy brown` = "#FA9E4D")
custom_col <- function(...) {
  cols <- c(...)
  if (is.null(cols))
    return (drsimonj_colors)
  custom_colors[cols]
}
custom_palettes <- list(
  `main`  = custom_col("green", "yellow", "dark red"),
  `cool`  = custom_col("green", "yellow","orange"),
  `hot`   = custom_col("yellow", "dark red", "sandy brown"),
  `simple` = custom_col("yellow", "dark red"),
  `reallysimple` = custom_col("white", "dark red"),
  `mixed` = custom_col("green", "yellow", "orange", "bright red", "dark red"),
  `distinct` = custom_col("dark red", "bright red", "light yellow", "green","grey"),
  `distinct1` = custom_col("dark red", "bright red",  "green","light yellow","grey"), 
  `bar` = custom_col("green", "grey", "yellow", "orange", "bright red", "dark red"),
  `treatment` = custom_col("dark red","green", "space sparkle", "orange", "violet","yellow","sandy brown","bright red"),
  `exception` = custom_col("dark red", "yellow"),
  `agegroup` = custom_col("bright red", "orange", "yellow", "grey")
)
custom_pal <- function(palette = "main", reverse = FALSE, ...) {
  pal <- custom_palettes[[palette]]
  if (reverse) pal <- rev(pal)
  colorRampPalette(pal, ...)
}
scale_color_custom <- function(palette = "mixed", discrete = TRUE, reverse = FALSE, ...) {
  pal <- custom_pal(palette = palette, reverse = reverse)
  if (discrete) {
    discrete_scale("colour", paste0("custom_", palette), palette = pal, ...)
  } else {
    scale_color_gradientn(colours = pal(256), ...)
  }
}
scale_fill_custom <- function(palette = "mixed", discrete = TRUE, reverse = FALSE, ...) {
  pal <- custom_pal(palette = palette, reverse = reverse)
  if (discrete) {
    discrete_scale("fill", paste0("custom_", palette), palette = pal, ...)
  } else {
    scale_fill_gradientn(colours = pal(256), ...)
  }
}

```

```{r fig2}
exceptions <- filter(p1, REC_EXCEPTION == 1)
d1 <- p1 %>%
  mutate(
    type = case_when(
      REC_EXCEPTION == 0 ~ 0,
      REC_EXCEPTION == 1 ~ 1
    )
  )
exceptions <- exceptions %>% 
  mutate(
    type = case_when(
      REC_EXCEPTION == 1 ~ 2
    )
  )
d1 <- rbind(d1, exceptions)
d1 <- d1 %>% 
  mutate(
    newscore = case_when(
      type == 0 ~ SCORE,
      type == 1 ~ SCORE,
      type == 2 ~ LAB_SCORE
    ),
    except_type = case_when(
      type == 0 ~ "Allocation/Laboratory Score of Patients without Exceptions",
      type == 1 ~ "Allocation Score of Patients with Exceptions",
      type == 2 ~ "Laboratory Score of Patients with Exceptions"
    )
  )
d1$except_type <- as.factor(d1$except_type)

ggplot(data = d1, aes(x = newscore)) +
  geom_histogram(aes(fill = except_type, alpha = except_type == "Allocation/Laboratory Score of Patients without Exceptions"), color = "black",
                 position = "identity", bins = 60) +
  theme_apa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top") +
  scale_fill_discrete(name = "") +
  labs(x = "MELD/PELD Score", y = "Number of Candidates Listed") + 
  scale_x_continuous(breaks = c(-20, -10, 0, 10, 20, 30, 40, 50, 60, 70)) +
  scale_alpha_manual(values = c(1, 0.3), guide = "none") +
  guides(fill = guide_legend(nrow = 3))
```


```{r fig3}
matched_by_allocation$REC_EXCEPTION <- as.factor(matched_by_allocation$REC_EXCEPTION)
matched_by_allocation$removaltype <- as.factor(matched_by_allocation$removaltype)
levels(matched_by_allocation$removaltype) <- c("On List", "Death", "Transplant")
levels(matched_by_allocation$REC_EXCEPTION) <- c("No Exception, Matched by Allocation Score", "Exception")

newfx <- survfit2(Surv(TIME_TO_REMOVAL, removaltype) ~ REC_EXCEPTION, data = matched_by_allocation) %>%
  ggcuminc(
    outcome = c("Death", "Transplant")) + 
  palette = c("#660000", "#e49f62") +
  add_confidence_interval() +
  coord_cartesian(xlim = c(0, 180)) +
  scale_x_continuous(breaks = c(0, 60, 120, 180)) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent, 
    expand = c(0.01, 0)
  ) +
  xlab("Days after Listing") +
  theme_classic() +
  guides(linetype=guide_legend(nrow=2, byrow=TRUE)) +
  guides(color=guide_legend(nrow=2, byrow=TRUE)) +
  ggtitle("A") +
  theme(plot.title = element_text(size = 18),
        legend.title = element_blank(),
        legend.position = "bottom") +
  annotate("text", x = 0, y = 0.95, hjust = 0,
           label = paste0(
             "Death p: ", 
             ifelse(cif$Tests[1, 2] < .001, 
                    "< 0.001", 
                    round(cif$Tests[1, 2], 3)))) + 
  annotate("text", x = 0, y = 0.92, hjust = 0,
           label = paste0(
             "Transplant p: ", 
             ifelse(cif$Tests[2, 2] < .001, 
                    "< 0.001", 
                    round(cif$Tests[2, 2], 2)))) 

matched_by_laboratory$REC_EXCEPTION <- as.factor(matched_by_laboratory$REC_EXCEPTION)
matched_by_laboratory$removaltype <- as.factor(matched_by_laboratory$removaltype)
levels(matched_by_laboratory$removaltype) <- c("On List", "Death", "Transplant")
levels(matched_by_laboratory$REC_EXCEPTION) <- c("No Exception, Matched by Laboratory Score", "Exception")

newfx1 <- survfit2(Surv(TIME_TO_REMOVAL, removaltype) ~ REC_EXCEPTION, data = matched_by_laboratory) %>%
  ggcuminc(
    outcome = c("Death", "Transplant")) + 
  add_confidence_interval() +
  coord_cartesian(xlim = c(0, 180)) +
  scale_x_continuous(breaks = c(0, 60, 120, 180)) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent, 
    expand = c(0.01, 0)
  ) +
  xlab("Days after Listing") +
  theme_classic() +
  guides(linetype=guide_legend(nrow=2, byrow=TRUE)) +
  guides(color=guide_legend(nrow=2, byrow=TRUE)) +
  ggtitle("B") +
  theme(plot.title = element_text(size = 18),
        legend.title = element_blank(),
        legend.position = "bottom") +
  annotate("text", x = 0, y = 0.95, hjust = 0,
           label = paste0(
             "Death p: ", 
             ifelse(cif1$Tests[1, 2] < .001, 
                    "< 0.001", 
                    "0.16"))) + 
  annotate("text", x = 0, y = 0.92, hjust = 0,
           label = paste0(
             "Transplant p: ", 
             ifelse(cif1$Tests[2, 2] < .001, 
                    "< 0.001", 
                   "< 0.001"))) 

newfx + newfx1
```

Below is the code for performing Fine-Gray competing risks regression analysis to determine the sub-distribution hazard ratio for cumulative incidence of waitlist removal due to death and DDLT.
```{r fine_gray}

#sHR of removal due to DDLT in cohort matched by allocation score, with death treated as competing risk.
crrtx <- 
  crr(
    ftime = matched_by_allocation$TIME_TO_REMOVAL,
    fstatus = matched_by_allocation$removaltype,
    cov1 = matched_by_allocation[, c("REC_EXCEPTION")],
    failcode = 2,
    cencode = 0
  )

summary(crrtx, conf.int = 0.95)

#sHR of removal due to death in cohort matched by allocation score, with DDLT treated as competing risk.
crrtx1 <- 
  crr(
    ftime = matched_by_allocation$TIME_TO_REMOVAL,
    fstatus = matched_by_allocation$removaltype,
    cov1 = matched_by_allocation[, c("REC_EXCEPTION")],
    failcode = 1,
    cencode = 0
  )

summary(crrtx1, conf.int = 0.95)

#sHR of removal due to DDLT in cohort matched by laboratory score, with death treated as competing risk.
crrtx2 <- 
  crr(
    ftime = matched_by_laboratory$TIME_TO_REMOVAL,
    fstatus = matched_by_laboratory$removaltype,
    cov1 = matched_by_laboratory[, c("REC_EXCEPTION")],
    failcode = 2,
    cencode = 0
  )

summary(crrtx2, conf.int = 0.95)

#sHR of removal due to death in cohort matched by laboratory score, with DDLT treated as competing risk.
crrtx3 <- 
  crr(
    ftime = matched_by_allocation$TIME_TO_REMOVAL,
    fstatus = matched_by_allocation$removaltype,
    cov1 = matched_by_allocation[, c("REC_EXCEPTION")],
    failcode = 1,
    cencode = 0
  )

summary(crrtx3, conf.int = 0.95)

```

```{r fig5}

ddlt <- pedsexception %>% group_by(PX_ID) %>% filter(CAN_REM_CD == 4) %>% 
  arrange(CANHX_BEGIN_DT) %>% slice_tail(n = 1) %>%
  mutate(EXCEPTION_AT_TRANSPLANT = case_when(
    (CANHX_MPXCPT_STAT == 5 | CANHX_MPXCPT_STAT == 14 | CANHX_MPXCPT_STAT == 16 | CANHX_MPXCPT_STAT == 18 | CANHX_MPXCPT_STAT == 30) ~ 1,
    TRUE ~ 0)) %>% filter(!(EXCEPTION_AT_TRANSPLANT == 1 & CANHX_MPXCPT_DGN != 9))

hadexception <- filter(ddlt, EXCEPTION_AT_TRANSPLANT == 1) %>% 
  mutate(EXCEPTION_STATUS = case_when(
    EXCEPTION_AT_TRANSPLANT == 1 ~ "Active Exception at Time of Transplant"))
a <- ddlt %>% 
  mutate(EXCEPTION_STATUS = case_when(
    EXCEPTION_AT_TRANSPLANT == 0 | EXCEPTION_AT_TRANSPLANT == 1 ~ "Total"))
hadexception <- data_frame(hadexception)
a <- data_frame(a)
f5data <- rbind(hadexception, a)
finalset1 <- f5data %>%
  mutate(month = zoo::as.yearmon(CAN_LISTING_DT),
         status = factor(EXCEPTION_STATUS,
                         levels = c("Total", "Active Exception at Time of Transplant")))
by_month1 <- finalset1 %>%
  group_by(POLICY_COHORT, month) %>%
  count(status) %>%
  mutate(total_per_month = sum(n)) %>%
  ungroup() %>%
  mutate(month1 = factor(month),
         percentage_per_month = 100*n/total_per_month) 
policy_switch <- which(levels(by_month$month1) == "May 2019")
pre_policy_start <- which(levels(by_month$month1) == "Jun 2016")
pre_policy_end <- which(levels(by_month$month1) == "Sep 2018")
post_policy_start <- which(levels(by_month$month1) == "Jun 2019")
post_policy_end <- which(levels(by_month$month1) == "Sep 2021")

fig5 <- ggplot(by_month1, aes(x = month1, 
                              y = n, 
                              color = status, 
                              group = status)) +
  geom_line() + 
  geom_point() +
  geom_vline(aes(xintercept = policy_switch, linetype = "Policy Switch")) + 
  scale_color_custom(name = "Exception Status at Transplant", palette = "simple") +
  labs(
    x = "Month of Listing",
    y = "Number of Transplant Recipients",
    linetype = "",
    color = "status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.title.x = element_text(margin = margin(t = 15, r = 0, b = 0, l = 0)),
        legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.margin = margin(t=0, r=0, b=0, l=0, unit = "mm"),
        legend.direction = "vertical",
        legend.title.align = 0,
        panel.background = element_rect(fill = "grey99"),
        plot.background = element_rect(fill = "white"),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black")
  ) +
  scale_linetype_manual(values = c("dashed", "dotted", "solid")) +
  scale_fill_custom(palette = "simple") +
  guides(colour = guide_legend(nrow = 1)) +
  annotate("text", x = which(levels(by_month$month1) == "Aug 2017"), y = 45, label= "Pre-Policy Cohort") +
  annotate("text", x = which(levels(by_month$month1) == "Aug 2020"), y = 45, label= "Post-Policy Cohort") +
  annotate("rect",
           xmin = which(levels(by_month$month1) == "Jun 2016"),           
           xmax = which(levels(by_month$month1) == "Sep 2018"),
           ymin = 0, 
           ymax = 50,  
           alpha = 0.15) +
  annotate("rect",
           xmin = which(levels(by_month$month1) == "Jun 2019"),            
           xmax = which(levels(by_month$month1) == "Sep 2021"),            
           ymin = 0, 
           ymax = 50, 
           alpha = 0.15) 



f6data <- ddlt %>% mutate(EXCEPTION_STATUS = case_when(
  EXCEPTION_AT_TRANSPLANT == 0 ~ "No Exception",
  EXCEPTION_AT_TRANSPLANT == 1 ~ "Active Exception at Time of Transplant"
))

finalset <- f6data %>%
  mutate(month = zoo::as.yearmon(CAN_LISTING_DT),
         status = factor(EXCEPTION_STATUS,
                         levels = c("No Exception", "Active Exception at Time of Transplant")))
by_month <- finalset %>%
  group_by(POLICY_COHORT, month) %>%
  count(status) %>%
  mutate(total_per_month = sum(n)) %>%
  ungroup() %>%
  mutate(month1 = factor(month),
         percentage_per_month = n*100/total_per_month) 
policy_switch <- which(levels(by_month$month1) == "May 2019")
pre_policy_start <- which(levels(by_month$month1) == "Jun 2016")
pre_policy_end <- which(levels(by_month$month1) == "Sep 2018")
post_policy_start <- which(levels(by_month$month1) == "Jun 2019")
post_policy_end <- which(levels(by_month$month1) == "Sep 2021")

fig6 <- ggplot(by_month, aes(x = month1, 
                             y = percentage_per_month, 
                             fill = status, 
                             group = status)) +
  geom_col(alpha = 0.75) +
  geom_vline(aes(xintercept = policy_switch, linetype = "Policy Switch")) + 
  labs(
    x = "",
    y = "% Exceptions at Transplant",
    linetype = "",
    fill = "status") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.title.x = element_text(margin = margin(t = 15, r = 0, b = 0, l = 0)),
        legend.position = "none",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.margin = margin(t=0, r=0, b=0, l=0, unit = "mm"),
        legend.direction = "vertical",
        legend.title.align = 0,
        panel.background = element_rect(fill = "grey99"),
        plot.background = element_rect(fill = "white"),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black")
  ) +
  scale_linetype_manual(values = c("dashed", "dotted", "solid")) +
  scale_fill_custom(name = "Status",
                    breaks = c("Active Exception at Time of Transplant"),
                    palette = "reallysimple") +
  guides(colour = guide_legend(nrow = 1)) +
  annotate("text", x = which(levels(by_month$month1) == "Aug 2017"), y = 95, label= "Pre-Policy Cohort") +
  annotate("text", x = which(levels(by_month$month1) == "Aug 2020"), y = 95, label= "Post-Policy Cohort") +
  annotate("rect",
           xmin = pre_policy_start,           
           xmax = pre_policy_end,
           ymin = 0, 
           ymax = 100,  
           alpha = 0.15) +
  annotate("rect",
           xmin = post_policy_start,            
           xmax = post_policy_end,            
           ymin = 0, 
           ymax = 100, 
           alpha = 0.15) 

plot_grid(fig5, fig6,
          ncol = 1, nrow = 2,
          labels = c("A", "B"))

```

